---
title: "EDUC 452 PS4"
author: "###Jason Zhang###"
date: "`r Sys.time()`"
output:
  html_document:
    df_print: paged
  pdf_document:
    number_sections: true
urlcolor: blue
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
# load packages
library("knitr")
library("tidyverse")

# use a nicer plotting theme theme
theme_set(theme_classic())

# disable warnings about grouping
options(dplyr.summarise.inform = FALSE)
```

> This homework is due by **Friday, Jun. 6th, 8:00am**.
> Upload a html file to Canvas called `ps4.html` I know it’s a busy time of the year, so you can choose just one of the below to submit.
> Please also see the ‘independent’ assignment which will need to be submitted independently.\\

```{r}
library("knitr")
library("tidyverse")

theme_set(theme_classic())

options(dplyr.summarise.inform = FALSE)

library(haven)
library(ggplot2)

df <- read.csv("els_cleaned.csv")

df_clean <- df[complete.cases(df[c("BYSES1", "F3ICNTRL", "F3ERN2011")]), ]

df_clean$private <- ifelse(df_clean$F3ICNTRL == 2, 1, 0)

std <- function(x) (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
df_clean$ses_std <- std(df_clean$BYSES1)
df_clean$earn_log <- log(df_clean$F3ERN2011 + 1)

print("=== Real Data Analysis ===")
print(paste("N students:", nrow(df_clean)))
print(paste("Proportion attending private:", mean(df_clean$private)))
print(paste("Mean SES by institution type:"))
print(aggregate(ses_std ~ private, df_clean, mean))

real_model <- lm(earn_log ~ ses_std * private, data = df_clean)
print(summary(real_model))

beta_0 <- coef(real_model)[1]  # Intercept
beta_ses <- coef(real_model)[2]  # SES effect
beta_private <- coef(real_model)[3]  # Private school effect
beta_interaction <- coef(real_model)[4]  # SES × Private interaction
sigma <- sigma(real_model)  # Residual SD

simulate_data <- function(n = nrow(df_clean), 
                         ability_effect = 0.3,  # Hidden confounder
                         selection_strength = 0.5) {  # How much ability affects school choice
  
  ses <- rnorm(n, 0, 1)  # Standardized SES
  ability <- 0.4 * ses + rnorm(n, 0, 0.8)  # Ability correlated with SES
  
  private_prob <- plogis(-1.5 + 0.8 * ses + selection_strength * ability)
  private <- rbinom(n, 1, private_prob)
  
  earnings_log <- beta_0 + 
                  beta_ses * ses + 
                  beta_private * private + 
                  beta_interaction * ses * private +
                  ability_effect * ability +  # Hidden confounder effect
                  rnorm(n, 0, sigma)
  
  data.frame(
    ses = ses,
    private = private,
    ability = ability,
    earnings_log = earnings_log
  )
}

n_sims <- 500
results <- data.frame(
  sim = 1:n_sims,
  true_effect = NA,
  naive_effect = NA,
  ses_adjusted = NA,
  full_model = NA
)

set.seed(123)
for (i in 1:n_sims) {
  sim_data <- simulate_data()
  
  true_model <- lm(earnings_log ~ ses * private + ability, data = sim_data)
  results$true_effect[i] <- coef(true_model)["private"]
  
  # DAM 1: Naive comparison (no controls)
  naive_model <- lm(earnings_log ~ private, data = sim_data)
  results$naive_effect[i] <- coef(naive_model)["private"]
  
  # DAM 2: Control for SES only
  ses_model <- lm(earnings_log ~ ses + private, data = sim_data)
  results$ses_adjusted[i] <- coef(ses_model)["private"]
  
  # DAM 3: Full model with interaction (but still missing ability)
  full_model <- lm(earnings_log ~ ses * private, data = sim_data)
  results$full_model[i] <- coef(full_model)["private"]
}

print("\n=== Simulation Results ===")
print("Average estimates across simulations:")
print(colMeans(results[, -1]))

print("\nBias (relative to true effect):")
bias <- colMeans(results[, -1]) - mean(results$true_effect)
print(bias[-1])

library(tidyr)
results_long <- results %>%
  pivot_longer(cols = -sim, names_to = "model", values_to = "estimate")

p <- ggplot(results_long, aes(x = estimate, fill = model)) +
  geom_density(alpha = 0.6) +
  geom_vline(xintercept = mean(results$true_effect), 
             linetype = "dashed", size = 1) +
  labs(
    title = "Distribution of Private School Effect Estimates",
    subtitle = "Dashed line = True causal effect",
    x = "Estimated Effect on Log Earnings",
    y = "Density"
  ) +
  scale_fill_manual(
    values = c("true_effect" = "black", 
               "naive_effect" = "red", 
               "ses_adjusted" = "blue", 
               "full_model" = "green"),
    labels = c("True Effect", "Naive", "SES Adjusted", "Full Model")
  ) +
  theme_minimal()

print(p)

# ggsave("simulation_results.png", p, width = 8, height = 6)

print("\n=== Sensitivity Analysis ===")
selection_strengths <- seq(0, 1, 0.2)
sensitivity_results <- data.frame(
  selection = selection_strengths,
  bias_naive = NA,
  bias_ses = NA,
  bias_full = NA
)

for (j in 1:length(selection_strengths)) {
  temp_results <- replicate(100, {
    sim_data <- simulate_data(selection_strength = selection_strengths[j])
    
    true_model <- lm(earnings_log ~ ses * private + ability, data = sim_data)
    true_effect <- coef(true_model)["private"]
    
    naive <- coef(lm(earnings_log ~ private, data = sim_data))["private"]
    ses_adj <- coef(lm(earnings_log ~ ses + private, data = sim_data))["private"]
    full <- coef(lm(earnings_log ~ ses * private, data = sim_data))["private"]
    
    c(naive - true_effect, ses_adj - true_effect, full - true_effect)
  })
  
  sensitivity_results[j, -1] <- rowMeans(temp_results)
}

print("Bias by selection strength:")
print(sensitivity_results)

sens_long <- sensitivity_results %>%
  pivot_longer(cols = -selection, names_to = "model", values_to = "bias")

p2 <- ggplot(sens_long, aes(x = selection, y = bias, color = model)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Selection Bias Sensitivity Analysis",
    x = "Selection Strength (ability → private school)",
    y = "Bias in Estimated Effect"
  ) +
  scale_color_manual(
    values = c("bias_naive" = "red", "bias_ses" = "blue", "bias_full" = "green"),
    labels = c("Naive", "SES Adjusted", "Full Model")
  ) +
  theme_minimal()

print(p2)
ggsave("sensitivity_analysis.png", p2, width = 8, height = 6)
```

That's all!
Thank you for engaging the 4 problem sets :) Have a wonderful summer!

# Session info

Information about this R session including which version of R was used, and what packages were loaded.

```{r sessinfo}
sessionInfo()
```
